---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(here)

# Load the cleaned speed data
data <- read_csv(here::here("data/02-analysis_data/analysis_speed_data.csv"))

# Ensure the data is ready for analysis
data <- data %>%
  mutate(
    month = as.Date(month)  # Ensure the 'month' column is in Date format
  )
```


# Introduction



# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Traffic Speed Dataset

The traffic speed dataset contains information collected from multiple locations and months, detailing percentile speeds, total traffic volume, and counts of vehicles traveling within specific speed ranges. The dataset aggregates these metrics monthly, offering insights into traffic behavior and variations over time.

This dataset includes:
- Percentile speeds (5th, 10th, ... 95th) representing speed thresholds exceeded by corresponding proportions of vehicles.
- Traffic volume as the total number of vehicles observed monthly.
- Speed bins (e.g., 0-4 km/h, 5-9 km/h) for detailed breakdowns of vehicle counts across speed ranges.
- The table below, @tbl-speed-preview, showcases a preview of the cleaned dataset with selected columns to highlight key features.

```{r}
#| label: tbl-speed-preview
#| tbl-cap: Preview of cleaned traffic speed dataset.
#| echo: false
#| tbl-pos: H

data |> 
  select(month, pct_05, pct_50, pct_95, spd_30, spd_50, spd_100_and_above, volume) |> 
  head(10) |> 
  kable(
    col.names = c("Month", "5th Percentile", "Median (50th Percentile)", "95th Percentile", 
                  "Vehicles (30-34 km/h)", "Vehicles (50-54 km/h)", "Vehicles (>100 km/h)", "Total Volume"),
    booktabs = TRUE
  ) |> 
  column_spec(1, width = "2.5cm") |> 
  column_spec(2:8, width = "2.0cm")

```

## Random Sample from the Dataset
@tbl-random-sample provides a random sample of 10 observations, offering an unbiased snapshot of key metrics such as speed percentiles, total traffic volume, and counts of vehicles exceeding 100 km/h. This selection illustrates the dataset's diversity without emphasizing specific patterns.

```{r}
#| label: tbl-random-sample
#| tbl-cap: Random sample of traffic speed dataset.
#| echo: false
#| tbl-pos: H

set.seed(304)  # For reproducibility
data |> 
  sample_n(10) |> 
  select(month, pct_05, pct_50, pct_95, volume, spd_100_and_above) |> 
  kable(
    col.names = c("Month", "5th Percentile", "Median (50th Percentile)", "95th Percentile", "Total Volume", "Vehicles (>100 km/h)"),
    booktabs = TRUE
  ) |> 
  column_spec(1, width = "2.5cm") |> 
  column_spec(2:6, width = "2.0cm")

```

## Extreme Speeds Dataset
Vehicles traveling at extreme speeds (over 100 km/h) represent a critical factor in understanding traffic safety and violations. The table below, @tbl-extreme-speeds, highlights the top 5 months with the highest counts of vehicles exceeding 100 km/h. This provides insights into when extreme speeds are most prevalent.

```{r}
#| label: tbl-extreme-speeds
#| tbl-cap: Top 5 months with the highest counts of vehicles exceeding 100 km/h.
#| echo: false
#| tbl-pos: H

data |> 
  arrange(desc(spd_100_and_above)) |> 
  head(5) |> 
  select(month, volume, spd_100_and_above, pct_95) |> 
  kable(
    col.names = c("Month", "Total Volume", "Vehicles (>100 km/h)", "95th Percentile Speed"),
    booktabs = TRUE
  ) |> 
  column_spec(1, width = "2.5cm") |> 
  column_spec(2:4, width = "2.0cm")

```

## Proportion of Vehicles at Moderate Speeds
Another key insight from this dataset is the proportion of vehicles traveling within a moderate speed range (50–70 km/h). This measure helps understand how typical driving speeds align with safe and efficient traffic flow.

The table below, @tbl-speed-proportion, highlights monthly proportions of vehicles in this speed range.

```{r}
#| label: tbl-speed-proportion
#| tbl-cap: Monthly proportions of vehicles traveling at moderate speeds (50–70 km/h).
#| echo: false
#| tbl-pos: H

moderate_speed <- data %>%
  mutate(month = lubridate::month(month, label = TRUE, abbr = TRUE)) %>%
  group_by(month) %>%
  summarise(
    moderate_speed_count = sum(spd_50 + spd_55 + spd_60 + spd_65 + spd_70, na.rm = TRUE),
    total_volume = sum(volume, na.rm = TRUE),
    proportion_moderate = round((moderate_speed_count / total_volume) * 100, 2)
  ) 

moderate_speed |> 
  kable(
    col.names = c("Month", "Moderate Speed Vehicles", "Total Volume", "Proportion (%)"), 
    booktabs = TRUE
  ) |> 
  column_spec(1, width = "2.5cm") |> 
  column_spec(2:4, width = "2.5cm")

```













```{r}
#| label: fig-extreme-speeds-vs-volume
#| fig-cap: Relationship between extreme speeds (over 100 km/h) and total traffic volume, categorized by pre-2020 and post-2020 time periods.
#| echo: false
#| warning: false
#| message: false

# Process the data for analysis
extreme_vs_volume <- data %>%
  group_by(month) %>%
  summarise(
    extreme_count = sum(spd_100_and_above, na.rm = TRUE),
    total_volume = sum(volume, na.rm = TRUE)
  ) %>%
  mutate(period = ifelse(month < as.Date("2020-01-01"), "Pre-2020", "Post-2020"))

# Plot Extreme Speeds vs Total Traffic Volume by Time Period
ggplot(extreme_vs_volume, aes(x = total_volume, y = extreme_count, color = period)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Total Traffic Volume",
    y = "Extreme Speeds (Over 100 km/h)",
    color = "Time Period"
  ) +
  theme_minimal()


```

```{r}
#| label: fig-key-percentiles
#| fig-cap: Trends in key speed percentiles (5th, 50th, and 95th) and the spread between the 95th and 5th percentiles over time.
#| echo: false
#| warning: false
#| message: false
# Summarize Key Percentiles and Calculate Speed Spread
key_percentiles <- data %>%
  group_by(month) %>%
  summarise(
    pct_05 = mean(pct_05, na.rm = TRUE),
    pct_50 = mean(pct_50, na.rm = TRUE),
    pct_95 = mean(pct_95, na.rm = TRUE)
  ) %>%
  mutate(speed_spread = pct_95 - pct_05)  # Calculate spread

# Plot Key Percentiles and Spread
ggplot(key_percentiles, aes(x = month)) +
  geom_line(aes(y = pct_05, color = "5th Percentile")) +
  geom_line(aes(y = pct_50, color = "Median (50th Percentile)")) +
  geom_line(aes(y = pct_95, color = "95th Percentile")) +
  geom_line(aes(y = speed_spread, color = "Speed Spread"), linetype = "dashed") +
  labs(
    x = "Month",
    y = "Speed (km/h)",
    color = "Legend"
  ) +
  theme_minimal()

```


```{r}


```

Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.








# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


